Unbalanced Mallows Models for Optimizing Expensive Black-Box Permutation Problems

### 2021-02-02 paper/main.tex
more updates

% \author{Ekhiñe Irurozki\inst{1} \and Manuel López-Ibáñez\inst{2}\orcidID{0000-0001-9974-1295}}
% \institute{
%    Basque Center for Applied Mathematics\\
%    \email{eirurozki@bcamath.org}
%    \and
%    University of Málaga, Málaga, Spain\\
%    \email{manuel.lopez-ibanez@uma.es}
%  }
%    \renewcommand{\shortauthors}{B. Trovato et al.}
\begin{abstract}
%The abstract should briefly summarize the contents of the paper in
%150--250 words.
  Expensive black-box combinatorial optimization problems arise in practice
  when the objective function is evaluated by means of a simulator or a
  real-world experiment. Since each fitness evaluation is expensive in terms of
  time or resources, only a limited number of evaluations is possible,
  typically several orders of magnitude smaller than in non-expensive problems. In
  this scenario, classical optimization methods such as mixed-integer
  programming and local search are not useful.  In the continuous case,
  Bayesian optimization, in particular using Gaussian processes, has proven
  very effective under these conditions. Much less research is available in the
  combinatorial case. In this paper, we propose and analyze UMM, an
  estimation-of-distribution (EDA) algorithm based on a Mallows probabilistic
  model and unbalanced rank aggregation (uBorda). Experimental results on
  black-box versions of LOP, QAP and PFSP show that UMM is able to match, and
  sometimes surpass, the solutions obtained by CEGO, a Bayesian optimization
  algorithm for combinatorial
  optimization. %, specially when the budget of solutions evaluated approaches 400.
  Moreover, the computational complexity of UMM increases linearly with both
  the number of function evaluations and the permutation size.
%  an order of magnitude smaller than CEGO, which makes it specially useful when
  %both computation time and number of evaluations are of concern.
  \sloppy
  %\textcolor{red}{16 pages max, deadline: 1 November 2020}
\end{abstract}
%
% The code below should be generated by the tool at
% http://dl.acm.org/ccs.cfm
% Please copy and paste the code instead of the example below. 
%
\begin{CCSXML}
<ccs2012>
 <concept>
  <concept_id>10010520.10010553.10010562</concept_id>
  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010575.10010755</concept_id>
  <concept_desc>Computer systems organization~Redundancy</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010553.10010554</concept_id>
  <concept_desc>Computer systems organization~Robotics</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
 <concept>
  <concept_id>10003033.10003083.10003095</concept_id>
  <concept_desc>Networks~Network reliability</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
</ccs2012>  
\end{CCSXML}

@@

Permutations are defined as bijections of the set $[n]$ of integers onto itself. The set of all permutations of $n$ values is denoted as $S_n$ and has cardinality $n!$. We will use the one-line notation and denote permutations with Greek letters, $\sigma  = (\sigma(1), \ldots , \sigma(n))$. We denote the composition of $\sigma$ and $\pi$ as $\sigma\pi$ and the inverse of $\sigma$ as $\sigma^{-1}$, for which the relation $\sigma\sigma^{-1}=(1, 2, 3, \dotsc,n)$ always holds.

Permutation $\sigma$ represents a ranking of item:  $\sigma(i)=j$ denotes that item $i$ has rank $j$. The ordering associated with $\sigma$ is given by its inverse $\sigma^{-1}$, therefore,  $\sigma(i)=j \Leftrightarrow  \sigma^{-1}(j)=i$. This distinction is in order since statistical methodologies are devoted for rankings and optimizaiton literature for orderings and mixing both concepts is a very common mistake. We will illustrate in the experimental section the consequences of this confusion.

Kendall's-$\tau$ distance is the metric for rankings that counts the pairs of items ranked in different order in both permutations:

\begin{equation}
\begin{split}
d(\sigma _{1},\sigma_{2})=\left\lvert\Bigl\{(i,j):i<j \land \phantom{}\Bigl(\right.&\bigl(\sigma_{1}(i) < \sigma_{1}(j)\land \sigma_{2}(i)>\sigma_{2}(j)\bigr)\lor \phantom{} \\
 &\bigl(\sigma_{1}(i)>\sigma_{1}(j)\land \sigma_{2}(i)<\sigma_{2}(j)\bigr)\Bigr)\Bigr\}\Bigr\rvert\\
 \end{split}
 \label{eq:kendall_decomp}
\end{equation}

>>>>

% \author{Ekhiñe Irurozki\inst{1} \and Manuel López-Ibáñez\inst{2}\orcidID{0000-0001-9974-1295}}
% \institute{
%    Basque Center for Applied Mathematics\\
%    \email{eirurozki@bcamath.org}
%    \and
%    University of Málaga, Málaga, Spain\\
%    \email{manuel.lopez-ibanez@uma.es}
%  }
%    \renewcommand{\shortauthors}{B. Trovato et al.}
\begin{abstract}
%The abstract should briefly summarize the contents of the paper in
%150--250 words.
  Expensive black-box combinatorial optimization problems arise in practice
  when the objective function is evaluated by means of a simulator or a
  real-world experiment. Since each fitness evaluation is expensive in terms of
  time or resources, only a limited number of evaluations is possible,
  typically several orders of magnitude smaller than in non-expensive problems. In
  this scenario, classical optimization methods such as mixed-integer
  programming and local search are not useful.  In the continuous case,
  Bayesian optimization, in particular using Gaussian processes, has proven
  very effective under these conditions. Much less research is available in the
  combinatorial case. In this paper, we propose and analyze UMM, an
  estimation-of-distribution (EDA) algorithm based on a Mallows probabilistic
  model and unbalanced rank aggregation (uBorda). Experimental results on
  black-box versions of LOP and PFSP show that UMM is able to match, and
  sometimes surpass, the solutions obtained by CEGO, a Bayesian optimization
  algorithm for combinatorial
  optimization. %, specially when the budget of solutions evaluated approaches 400.
  Moreover, the computational complexity of UMM increases linearly with both
  the number of function evaluations and the permutation size.
%  an order of magnitude smaller than CEGO, which makes it specially useful when
  %both computation time and number of evaluations are of concern.
  \sloppy
  %\textcolor{red}{16 pages max, deadline: 1 November 2020}
\end{abstract}
%
% The code below should be generated by the tool at
% http://dl.acm.org/ccs.cfm
% Please copy and paste the code instead of the example below. 
%
\begin{CCSXML}
<ccs2012>
 <concept>
  <concept_id>10010520.10010553.10010562</concept_id>
  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010575.10010755</concept_id>
  <concept_desc>Computer systems organization~Redundancy</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010553.10010554</concept_id>
  <concept_desc>Computer systems organization~Robotics</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
 <concept>
  <concept_id>10003033.10003083.10003095</concept_id>
  <concept_desc>Networks~Network reliability</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
</ccs2012>  
\end{CCSXML}

@@

Permutations are defined as bijections of the set $[n]$ of integers onto itself. The set of all permutations of $n$ values is denoted as $S_n$ and has cardinality $n!$. We will use the one-line notation\MANUEL{what is one-line?} and denote permutations with Greek letters, $\sigma  = (\sigma(1), \ldots , \sigma(n))$. We denote the composition of $\sigma$ and $\pi$ as $\sigma\pi = (\sigma(\pi(1)), \ldots, \sigma(\pi(n)))$\MANUEL{Is this correct?} and the inverse of $\sigma$ as $\sigma^{-1}$, for which the relation $\sigma\sigma^{-1}=(1, 2, 3, \dotsc,n)$ always holds.

\manuel{P}{Throughout this paper and for consistency, p}ermutation $\sigma$ represents a \emph{ranking}:  $\sigma(i)=j$ denotes that item $i$ has rank $j$. The \emph{ordering} associated with $\sigma$ is given by its inverse $\sigma^{-1}$, therefore,  $\sigma(i)=j \Leftrightarrow  \sigma^{-1}(j)=i$. This distinction is in order since statistical methodologies \manuel{are devoted for}{typically assume} rankings \manuel{and}{whereas the} optimization literature \manuel{for}{often assumes} orderings. Mixing both concepts is a very common confusion with grave consequences, as we will illustrate in the experimental section.

Kendall's-$\tau$ distance is the metric for rankings that counts the pairs of items ranked in different order in both permutations:
%
\begin{equation}
\begin{split}
d(\sigma _{1},\sigma_{2})=\left\lvert\Bigl\{(i,j):i<j \land \phantom{}\Bigl(\right.&\bigl(\sigma_{1}(i) < \sigma_{1}(j)\land \sigma_{2}(i)>\sigma_{2}(j)\bigr)\lor \phantom{} \\
 &\bigl(\sigma_{1}(i)>\sigma_{1}(j)\land \sigma_{2}(i)<\sigma_{2}(j)\bigr)\Bigr)\Bigr\}\Bigr\rvert\\
 \end{split}
 \label{eq:kendall_decomp}
\end{equation}
